---
title: "***t*-Test Assumptions**"
subtitle: "**STA4173: Biostatistics** <br> Spring 2025"
format: 
  revealjs: 
    code-overflow: wrap
    df-print: paged
    embed-resources: true
    slide-number: true
    width: 1600
    height: 900
    html-math-method: katex
    theme:
      - default
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: source
---

## Introduction: Assumptions

- We have now learned one- and two-sample *t*-tests.

- Recall, when we have two samples, they can be independent samples or dependent samples.

    - Independent samples: two-sample *t*-test
    
    - Dependent samples: paired *t*-test (one-sample *t*-test on difference)
    
- Today we will discuss how to assess the assumptions on *t*-tests.

## Normality Assumption: Set Up

- All *t*-tests assume approximate normality of the data.

    - In the case of one-sample *t*-tests, the measure of interest must somewhat follow a normal distribution.

    - In the case of two-sample *t*-tests, the measure of interest *in each group* must somewhat follow a normal distribution.
    
- Note that a paired *t*-test is technically a one-sample *t*-test, so we will examine normality of the difference. 

## Normality Assumption: Set Up

- There are formal tests for normality ([see article here](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3693611/)), however, we will not use them.

    - Tests for normality are not well-endorsed by statisticians.
    
- Instead, we will assess normality using a quantile-quantile (q-q) plot.

    - This is a scatterplot that will form a 45&deg; line if the assumed distribution is correct.
    
    - [Here is more information about q-q plots](https://data.library.virginia.edu/understanding-q-q-plots/).

- We will create q-q plots for:

    - The measurements in the case of the one-sample *t*-test.

    - The measurements from each group in the case of the two-sample *t*-test.
    
    - The difference between the groups in the case of the paired *t*-test.
    
## Normality Assumption: R Syntax

- We will assess the normality assumption graphically using a q-q plot
    
- A package was written by a former student, [`classpackage`](https://github.com/ieb2/classpackage).

    - If you **are** working on the server, the package is already installed.

    - If you are **not** working on the server, please ask me for the code needed to install.
    
```{r}
#| echo: false
#| eval: false
# install.packages("devtools") - use this if R tells you it can't find install_github()
devtools::install_github("ieb2/classpackage", force = TRUE)
```

## Normality Assumption: Independent Data - R Syntax

- Once installed, we call the package,

```{r}
library(classpackage)
```

- While there are several functions in this package, we are currently interested in the `independent_qq_plot()` function.

```{r}
#| eval: false
dataset_name %>% independent_qq_plot(variable = "continuous variable",
                                     grouping_variable = "grouping variable")
```

- This will provide the the q-q plot for the two-sample *t*-test (i.e., for independent data).

## Normality Assumption: Independent Data - Example

- Recall the penguin example for the two-sample *t*-test.

    - Is the body mass different for males and females?

```{r}
penguins <- palmerpenguins::penguins
head(penguins, n=3)
```

- Requesting the q-q plot,

```{r}
#| eval: false
penguins %>% independent_qq_plot(variable = "body_mass_g",
                                 grouping_variable = "sex")
```

## Normality Assumption: Independent Data - Example

<center>
```{r}
#| echo: false
penguins %>% na.omit(sex) %>% independent_qq_plot(variable = "body_mass_g",
                                 grouping_variable = "sex")
```
</center>


## Normality Assumption: Dependent Data - R Syntax

- While there are several functions in the `classpackage` package, we are now interested in the `dependent_qq_plot()` function.

```{r}
#| eval: false
wide_data %>% dependent_qq_plot(variable = "Display Name of Continuous Variable",
                                grouping_variable = " ", # do not edit this line
                                first_group = "first_variable", # first column for comparison
                                second_group = "second_variable") # second column for comparison
```

- This will provide the the q-q plot for the paired *t*-test (i.e., for dependent data).

## Normality Assumption: Repair Estimates

- Recall the repair estimate example for the dependent *t*-test.

```{r}
garage <- tibble(g1 = c(17.6, 20.2, 19.5, 11.3, 13.0, 
                        16.3, 15.3, 16.2, 12.2, 14.8,
                        21.3, 22.1, 16.9, 17.6, 18.4), 
                 g2 = c(17.3, 19.1, 18.4, 11.5, 12.7, 
                        15.8, 14.9, 15.3, 12.0, 14.2, 
                        21.0, 21.0, 16.1, 16.7, 17.5))
```

- Requesting the q-q plot,

```{r}
#| eval: false
garage %>% dependent_qq_plot(variable = "estimate",
                             grouping_variable = "garage",
                             first_group = "g1",
                             second_group = "g2")
```

## Normality Assumption: Repair Estimates

<center>
```{r}
#| echo: false
garage %>% dependent_qq_plot(variable = "estimate",
                             grouping_variable = "garage",
                             first_group = "g1",
                             second_group = "g2")
```
</center>


## Wrap Up

- **Important note!!** 

    - *I do not expect you to agree with my assessment of q-q plots!* 
    - What I do expect is that you know what to do after making your assessment.

- Next up: 

    - What happens if we do not meet the assumption for a *t*-test....?





























