---
title: "**Wilcoxon Signed Rank**"
subtitle: "**STA4173: Biostatistics** <br> Spring 2025"
format: 
  revealjs: 
    code-overflow: wrap
    df-print: paged
    embed-resources: true
    slide-number: true
    width: 1600
    height: 900
    html-math-method: katex
    theme:
      - default
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: source
---

## Introduction

- Today we have discussed that we turn to nonparametric tests when we do not meet distributional assumptions for *t*-tests.

- If we do not meet the normality assumption for the paired *t*-test, we turn to the Wilcoxon signed rank.

- Like in the dependent *t*-test, we will analyze the *difference* between two values.

- Like in the Wilcoxon rank sum, we will be analyzing ranks.

## Wilcoxon Signed Rank

- Before ranking, we will find the difference between the paired observations and eliminate any 0 differences.

    - Note 1: elimniating 0 differences is the big difference between the other tests! 
    
    - Note 2: because we are eliminating 0 differences, this means that our sample size will update to the number of pairs with a non-0 difference.
    
- When ranking, we the differences are ranked based on the absolute value of the difference.

- We also keep the sign of the difference.

    - We will have positive ranks and negative ranks.

| *X* | *Y* | *D* | \|*D*\| | Rank |
|-----|-----|-----|---------|------|
| 5   | 8   | -3  | 3       | $-$ 1.5 |
| 8   | 5   | 3   | 3       | $+$ 1.5  |
| 4   | 4   | 0   | 0       | ---------   |

## Wilcoxon Signed Rank{.smaller}

**Hypotheses**

- $H_0:$ $M_d = M_0$ | $H_0:$ $M_d \le M_0$ | $H_0:$ $M_d \ge M_0$
- $H_1:$ $M_d \ne M_0$ | $H_1:$ $M_d > M_0$ | $H_1:$ $M_d < M_0$

**Test Statistic & *p*-Value**

- $T_0 = \min(T+,|T_-|)$ if two-tailed, $T_0 = T_+$ if left-tailed, and $T_0 = |T_-|$ if right-tailed.
- $p =$ (calculated by R :))

**Rejection Region**

- Reject $H_0$ if $p < \alpha$.

**Conclusion/Interpretation**

- [Reject or fail to reject] $H_0$.

- There [is or is not] sufficient evidence to suggest [alternative hypothesis in words].

## Wilcoxon Signed Rank

-   We will again use the [`wilcox.test()`](https://www.rdocumentation.org/packages/stats/versions/3.6.1/topics/wilcox.test) function to perform the test,

```{r}
#| eval: false
wilcox.test(dataset$variable1, dataset$variable2,
       alternative = "alternative",
       mu = hypothesized_value,
       paired = TRUE,
       exact = FALSE)
```

## Wilcoxon Signed Rank

- A stock analyst believes the median number of shares traded in Walgreens Boots Alliance (WBA) stock is greater than that in McDonald’s (MCD). Test the analyst’s belief at the $\alpha=0.05$ level of significance.

```{r}
#| echo: false
library(tidyverse)
```

```{r}
stocks <- tibble(WBA = c(8.9, 6.3, 6.2, 7.2, 2.8, 3.3, 23.6, 
                         6.0, 15.6, 5.2, 6.3, 10.1, 4.0, 8.4),
                 MCD = c(8.5, 7.6, 8.3, 10.4, 2.5, 2.6, 3.5, 
                         4.7, 9.0, 6.0, 5.6, 5.0, 4.4, 5.6))
```

- Recall the R syntax,

```{r}
#| eval: false
wilcox.test(dataset$variable1, dataset$variable2,
       alternative = "alternative",
       mu = hypothesized_value,
       paired = TRUE,
       exact = FALSE)
```

## Wilcoxon Signed Rank

- From the problem statement: *A stock analyst believes the median number of shares traded in Walgreens Boots Alliance (WBA) stock is greater than that in McDonald’s (MCD).*

```{r}
wilcox.test(stocks$WBA, stocks$MCD,
       alternative = "greater",
       paired = TRUE, 
       exact = FALSE)
```

## Wilcoxon Signed Rank{.smaller}

- **Hypotheses**

    - $H_0: \ M_{\text{WBA}} \le M_{\text{MCD}}$ OR $M_d \le 0$, where $d = \text{WBA} - \text{MCD}$
    - $H_1: \ M_{\text{WBA}} > M_{\text{MCD}}$ OR $M_d > 0$
    
- **Test Statistic and *p*-Value**

    - $V_0 = 69$
    - $p = 0.158$
    
- **Rejection Region**

    - Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$.
    
- **Conclusion/Interpretation**

    - Fail to reject $H_0$.
    
    - There is not sufficient evidence to suggest that the median stock shares traded is greater for WBA than for MCD.

## **Wrap Up**

- Today we reviewed the Wilcoxon signed rank test.
    - Nonparametric alternative to the paired *t*-test.

- This completes Module 1.

- Next: comparing three or more groups.






























