---
title: "**Wilcoxon Rank Sum**"
subtitle: "**STA4173: Biostatistics** <br> Spring 2025"
format: 
  revealjs: 
    code-overflow: wrap
    df-print: paged
    embed-resources: true
    slide-number: true
    width: 1600
    height: 900
    html-math-method: katex
    theme:
      - default
execute:
  echo: true
  warning: false
  message: false
  error: true
editor: source
---

## Introduction

- We last discussed assumptions on *t*-tests

    - Dependent / paired *t*-test: normality
    
    - Independent two-sample *t*-test: normality and variance
  
- If we break the normality assumption, we must look to nonparametric methods.

```{r}
#| echo: false
library(tidyverse)
```

## Introduction

- The *t*-tests we have already learned are considered *parametric* methods.

    - There is a distributional assumption on the test.
    
- *Nonparametric* methods do not have distributional assumptions.

    - We typically transform the data to their ranks and then perform calculations.

- Why don't we always use nonparametric methods?

    - They are often less efficient: a larger sample size is required to achieve the same probability of a Type I error.
    
    - They discard useful information :(
    
## Ranking Data

- In the nonparametric tests we will be learning, the data will be ranked.

- Let us first consider a simple example, $$x: \ 1, 7, 10, 2, 6, 8$$

- Our first step is to reorder the data:$$x: \ 1, 2, 6, 7, 8, 10$$

- Then, we replace with the ranks:$$R: \ 1, 2, 3, 4, 5, 6$$

## Ranking Data

- What if all data values are not unique?

    - We will assign the average ranks.

- For example, $$x: \ 9, 8, 8, 0, 3, 4, 4, 8$$

- Let's reorder:$$x: \ 0, 3, 4, 4, 8, 8, 8, 9$$

- Rank ignoring ties:$$R: \ 1, 2, 3, 4, 5, 6, 7, 8$$

- Now, the final rank:$$R: \ 1, 2, 3.5, 3.5, 6, 6, 6, 8$$

## Wilcoxon Rank Sum{.smaller}

**Hypotheses**

- $H_0:$ $M_1 - M_2 = M_0$ | $H_0:$ $M_1 - M_2 \le M_0$ | $H_0:$ $M_1 - M_2 \ge M_0$
- $H_1:$ $M_1 - M_2 \ne M_0$ | $H_1:$ $M_1 - M_2 > M_0$ | $H_1:$ $M_1 - M_2 < M_0$

**Test Statistic & *p*-Value**

- $T = \sum R_{\text{sample 1}} - \frac{n_1(n_1+1)}{2}$
- $p =$ (calculated by R :))

**Rejection Region**

- Reject $H_0$ if $p < \alpha$.

**Conclusion/Interpretation**

- [Reject or fail to reject] $H_0$.

- There [is or is not] sufficient evidence to suggest [alternative hypothesis in words].

## Wilcoxon Rank Sum

-   We will use the [`wilcox.test()`](https://www.rdocumentation.org/packages/stats/versions/3.6.1/topics/wilcox.test) function to perform the test,

```{r}
#| eval: false
wilcox.test(continuous_variable ~ grouping_variable,
            data = dataset_name,
            alternative = "alternative",
            mu = hypothesized_value,
            exact = FALSE)
```

- Like before, R will use the group that is "first" in the grouping variable.

    - "First" is in terms of numeric or alphabetical.

## Wilcoxon Rank Sum

- When exposed to an infection, a person typically develops antibodies. The extent to which the antibodies respond can be measured by looking at a person’s titer, which is a measure of the number of antibodies present. The higher the titer is, the more antibodies that are present. 

- The following data represent the titers of 11 ill people and 11 healthy people exposed to the tularemia virus in Vermont. 

- Is the level of titer in the ill group greater than the level of titer in the healthy group? Use the $\alpha = 0.10$ level of significance.

```{r}
titer_levels <- tibble(level = c(640, 160, 1280, 320, 80, 640, 640, 160, 1280, 640, 160, 
                                  10, 320, 160, 160, 320, 320, 10, 320, 320, 80, 640),
                       group = c(rep("ill",11), rep("healthy",11)))
```

- Recall the R syntax,

```{r}
#| eval: false
wilcox.test(continuous_variable ~ grouping_variable,
            data = dataset_name,
            alternative = "alternative",
            mu = hypothesized_value,
            exact = FALSE)
```

## Wilcoxon Rank Sum

- Is the level of titer in the ill group greater than the level of titer in the healthy group?

```{r}
wilcox.test(level ~ group, 
            data = titer_levels,
            alternative = "less",
            exact = FALSE)
```

## Wilcoxon Rank Sum{.smaller}

**Hypotheses**

- $H_0: \ M_{\text{ill}} \le M_{\text{healthy}}$ 
- $H_1: \ M_{\text{ill}} > M_{\text{healthy}}$
    
**Test Statistic and *p*-Value**

- $W_0 = 35$
- $p = 0.047$
    
**Rejection Region**

- Reject $H_0$ if $p < \alpha$; $\alpha = 0.10$.

**Conclusion/Interpretation**

- Reject $H_0$.
    
- There is sufficient evidence to suggest that the level of titer in the ill group is greater than the level of titer in the healthy group.

## **Wrap Up**

- Today we reviewed the Wilcoxon rank sum test.
    - Nonparametric alternative to the two-sample *t*-test.

- Next lecture:
    - Wilcoxon signed rank



























